Potential for remotely sensing riverine gas exchange and carbon efflux from SWOT observations

### 2021-05-18 manuscript_v3.Rmd
first rewrite of methods using ustar model

##2 Methods
To build BIKER, we first develop a mathematical $k_{600}$ model that relies on no a priori knowledge of river hydraulics (section 2.1) and then we set that model as the core of a McFLI (section 2.2). Following method descriptions, we describe the validation (section 2.3) and the workflow for comparing estimated bulk carbon effluxes from a suite of average flow velocity models (section 2.4). A flowchart detailing the entire study is given as Figure 1.

![Figure 1. Flowchart of methodology used in this study. We begin by deriving a mathmatical model for k600 that uses only river width and slope to assign its parameters (section 2.1). Then, we implement this model within a McFLI ('BIKER') to estimate k600 solely from river width and water surface slope (section 2.2). Finally, we couple BIKER with in situ CO2 data to compare BIKER-estimated gas fluxes with established in situ methods (section 2.4). See section 2.3 for the validation setup.](outputs/flowchart.jpg)

###2.1 Developing a $k_{600}$ model for RS
@raymondScalingGasTransfer2012a first used river hydraulics and the theoretical relationship between *eD* and gas exchange to estimate $k_{600}$ from average flow velocity and channel slope. This extended earlier work predicting the dissolved oxygen reaeration coefficient in a similar manner [e.g. @oconnorMechanismReaerationNatural1958; @tsivoglouTracerMeasurementReaeration1976]. Later, @ulsethDistinctAirWater2019a built upon @raymondScalingGasTransfer2012a by including measurements from steeper rivers and finding that two distinct $k_{600}$ regimes exist in low and high energy rivers. More specifically, @ulsethDistinctAirWater2019a scale $k_{600}$ using *eD*, where the resulting model parameters are significantly different whether *eD* is high or low. The general form of their $k_{600}$ function is reprinted as equation 2, with the two energy regimes represented by subscript *r*. To date, upscaling studies have assumed uniform flow ($S_0=S_h$) in order to train their parameters across large datasets that only have information on $S_0$.

$$k_{600}=a_r(eD)^{b_r}  \mathbf{(2)}$$

While equation 2 as trained by @ulsethDistinctAirWater2019a dataset achieves the goal of this manuscript and provides a best-to-date predictive accuracy for $k_{600}$, it requires a priori knowledge of *eD* to assign its differential model parameters. Because $eD \approx gS_hV$ [@tsivoglouTracerMeasurementReaeration1976], we therefore need a priori knowledge of *V* to use equation 2 in a McFLI. So, we must rederive the @ulsethDistinctAirWater2019a $k_{600}$ model such that model parameters are solely assigned using the SWOT observables *W* and $S_h$.

Using the @tsivoglouTracerMeasurementReaeration1976 equation for *eD* for unsteady flow conditions (channel slope $S_0 \neq S_h$- see text S1 for this model derivation and why assuming non-uniform flow is convenient for an algorithm ingesting SWOT data) in conjunction with Manning’s equation to relate average flow velocity to channel geometry, we rewrite equation 2 as equation 3. In the far-right form of equation 3, we use the Manning’s formulation from @hagemannBAMBayesianAMHGManning2017a which formulates depth via river width and area (in order to use SWOT-observables). $k_{600}$ is now defined as a function of $S_h$, channel roughness *n* $[T/L^{\frac{1}{3}}]$, channel width *W* [L], and channel area $A_0+dA [L^2]$ (discretized via median channel area $A_0$ and the change in channel area *dA*). Note that we estimate *dA* by assuming a rectangular river channel so that $dA \approx \delta H*W$ and is also solely a function of *W* and $S_h$. Finally, equation 3 is also defined by the parameters $a_i$ and $b_i$, which are described next.

@@

$$k_{600} = a_i(eD)^{b_i} \approx a_i(gSV)^{b_i} \approx a_i\left[gS_h^{\frac{3}{2}}\frac{1}{n}(A_0+dA)^\frac{2}{3}W^\frac{-2}{3}\right]^{b_i} \mathbf{(3)}$$

As will be explained in section 2.2, the only observations in the BIKER formulation of equation 3 are *W* and $S_h$. So, we must assign $a_i$ and $b_i$ using just *W* and $S_h$. To do this, we use the basic premise that the equation 2 relationship is different for different river sizes *i*. We use *W* and *S* to define *i* and train different $k_{600}$ functions for rivers with the same *i*. We fit this model to the @ulsethDistinctAirWater2019a data (Figure 2) for five possible sets of $a_i$ and $b_i$ parameters, as defined for rivers 10-50m wide, 50-100m wide, 100+m wide, low energy rivers 0-10m wide (*S* < 0.05), and high energy rivers 0-10m wide (slope > 0.05). We find that narrow rivers (< 10m wide) are best modeled by further breaking them down into low energy (*S* < 0.05) and high energy (*S* > 0.05) regimes in order to meet the assumptions necessary to fit a linear regression model as proposed by @raymondScalingGasTransfer2012a and @ulsethDistinctAirWater2019a (specifically, normally distributed residuals of the predictions). Another implication of this extension of the @ulsethDistinctAirWater2019a model is that we now use *a* and *b* parameters that implicitly reflect the distinct geomorphology of different rivers.

![Figure 2: Theoretical basis for the RS-able extension of the Ulseth et al. (2019) k600 upscaling model. We use the differenial relationship between k600 and eD for different river sizes to define a 'rule-based' regression model. This model assigns differeniable model parameters using just river width and slope, following the form in equation 3. Data from Ulseth et al. (2019).](outputs/k600/theory_plot.jpg)

**these two paragraphs need to make it clear before you get to the end that you believe the relationship to be linear, but you will not limit yourself to raymond and/or ulseth. Rather, you wish to find any number of linear relationships that best fit the k600 data as statistically and hydraulicly defined. THIS MIGHT NEED TO BE A NON-LINEAR FUNCTION. THE OTHER PROBLEM IS THAT ALL WE ARE JUST HIGHLIGHTING THAT ALL UPSCALING RESEARCH HAS BASICALLY ASSUMED UNIFORM FLOW....**

Finally, we tested whether this model can reasonably estimate $k_{600}$ relative to the original @ulsethDistinctAirWater2019a model, with the goal of at least replicating their performance. We compared it against the various *eD*-based $k_{600}$ models from @raymondScalingGasTransfer2012a that are commonly used in the literature. This validation was done using a classic 80/20 split of the @ulsethDistinctAirWater2019a dataset, where 80% of the data was used to fit these models and 20% was used for independent testing of their predictive performance. This meant that the parameter values were slightly different than those published in both papers, but model structures were identical. This was done because the @raymondScalingGasTransfer2012a models were originally trained on a subset of this dataset and we deemed it fairer to compare model peformances using identical data.

###2.2 Developing BIKER
With a validated equation 3 in hand, we turn next to remotely sensing $k_{600}$. The approach used here is strongly informed by @hagemannBAMBayesianAMHGManning2017a’s McFLI algorithm for ungauged RSQ via Bayesian inference, further explored in more recent work [@brinkerhoffConstrainingRemoteRiver2020]. These papers conceptualize discharge as a Bayesian remote sensing problem, which we largely follow here to conceptualize $k_{600}$ as a Bayesian remote sensing problem that can be solved using SWOT data. Below, we translate @hagemannBAMBayesianAMHGManning2017a’s ideas from discharge to $k_{600}$.

SWOT will measure river width and height (and thus water surface slope $S_h$) at approximately 200m intervals in pre-defined mass-conserved river reaches (**What is the PDD?**; SWORD). Following @hagemannBAMBayesianAMHGManning2017a, the observations needed for Bayesian inference are SWOT measurements of *W* and $S_h$, and all other terms in equation 3 (*n*, $A_0$, and $k_{600}$) are treated as statistical parameters that cannot be directly observed from SWOT.

We therefore rework equation 3 into a Bayesian sampling model to have all of the ‘observations’ on the left-hand side sampled from all the unknown model parameters on the right-hand side (equation 4, where *x* are SWOT observables given $\Theta$, a set of statistical parameters). Note that in this formal likelihood specification both sides of equation 4 are log-transformed (natural log) to produce the normal sampling model presented here, however they are written in equation 4 without that transformation for succinctness. The $σ_{k600}^2$ parameter refers to the uncertainty inherent in equation 3's estimates. This will be explained in detail below.

$$f(x|\Theta) = \left[a_i^{-1}(gS_h^\frac{-3}{2}W^\frac{2}{3})^{b_i}\right] \sim normal\left(\left[k_{600}^{-1}\frac{1}{n}(A_0+dA)^\frac{2}{3})^{b_i}\right], \sigma^2_{k600}\right)  \mathbf{(4)}$$

Equation four necessitates that Bayesian priors be assigned to model parameters *n*, $A_0$, and $k_{600}$. Bayesian priors formalize the a priori estimates (and uncertainties) for the non-remotely-sensed parameters. To state that more intuitively for these physical quantities, priors represent our ‘prior river knowledge’ of what *n*, $A_0$, and $k_{600}$ probably are for some river given that they cannot be remotely sensed. All priors are formalized within the model as truncated normal distributions of the log-transformed terms such that $log(X) \sim N(\mu, \sigma^2)$ for $\lambda < X < \epsilon$, using prior hyperparameters mean ($\mu$), standard deviation ($\sigma$), and upper ($\epsilon$) and lower bounds ($\lambda$) for parameter *X*. In order to avoid relying on in situ information, we assign prior hyperparameters using SWOT data only. *n* and $A_0$ prior hyperparameters were assigned following the method developed by @brinkerhoffConstrainingRemoteRiver2020. They developed a set of river channel prior hyperparameters for McFLI algorithms that are entirely RS-able and reflect differential channel hydraulics as a function of river geomorphology. They used an extensive database of field measurements and machine learning to identify patterns that associate the river width with river geomorphic properties, so these non-remotely sensible properties may be assigned to rivers given existing remotely sensed data (namely, river width from optical imagery). With this method, they showed substantial improvement in the accuracy of ungauged RSQ estimates with no changes to the algorithms or observations. This leaves only $k_{600}$ hyperparameters to be defined. We assigned the $k_{600}$ prior hyperparameters using a simple slope regression model trained on the @ulsethDistinctAirWater2019a dataset to assign $\mu$ (equation 6). $σ^2$ was set to the standard error of equation 6: 1.023. λ and ε were set to -3 and 2.7, respectively.

$$\mu_{k600}=3.22+0.347ln(S_h) if ln(S_h) < -4.634   \mathbf{(6a)}$$
$$\mu_{k600}=6.85+1.13ln(S_h) if ln(S_h)>-4.634 \mathbf{(6b)}$$

Finally, we estimate $σ_{k600}^2$ using Monte Carlo (MC) methods to approximate total uncertainty from equation 3. Uncertainty in equation 3 can stem from 1) $a_i$ and $b_i$ parameter uncertainty, 2) error in Manning’s approximation of the average flow velocity, and 3) measurement error. Total equation 3 uncertainty is a function of all three uncertainity sources, so we need to propogate the three uncertanties through the algebra of equation 3. Here, we use MC simulations to do this. MC simulations repeately sample from parameter distributions defined by their individual expectations and uncertainities to produce a distribution of model estimates from which an overall uncertainity term can be extracted. Here, we run 5,000 unique MC simulations on 5,000 sets of field measurements of river channel hydraulics from @brinkerhoffReconcilingAtaStationAtManyStations2019 and use the median uncertainity term across those 5,000 distributions as $\sigma_{k600}$. Each MC simulation is itself 10,000 runs, sampling from the normal distributions for $a_i$, $b_i$, the log-transformed flow velocity, and $S_h$. Uncertainty for velocity followed @hagemannBAMBayesianAMHGManning2017a, who estimated ln-transformed Manning’s equation error in discharge to be 0.25 (we assume that the 0.25 value is due soley from Manning's equation and not measurements of cross-sectional area in order to use here as the velocity error). The slope-error model that we used is explained in detail in Section 2.3 and relies on a recent analysis of expected errors in SWOT $S_h$ measurements [@durandHowWillRadar2020a].

Finally, we repeat the above workflow three times under three different uncertainty scenarios to assess the overall influence of each uncertainity source on the equation 3 estimates. Uncertainty scenario 1 propogates errors in $a_i$ and $b_i$ only, uncertainty scenario 2 additionally propogates errors in log-transformed velocity, and uncertainty scenario 3 additionally propogates errors in $S_h$.

With the sampling model (equation 4), prior distributions, and the $σ_{k600}^2$ parameter described, a joint posterior distribution conditional on the SWOT data is obtained. To approximate this distribution, we use a Markov Chain Monte Carlo (MCMC) algorithm. Because it is written in the Stan probabilistic programming language, we use a Hamiltonian Monte Carlo which reduces computation time relative to other sampling algorithms [again following @hagemannBAMBayesianAMHGManning2017a].

###2.3 BIKER validation setup
We validated BIKER on 22 rivers using observed $V$ and observed $S_h$ First, we detail the 22 rivers, and then we detail how ‘observed’ $k_{600}$ is calculated.

Because SWOT has yet to launch, it is standard practice to benchmark SWOT-related algorithms on ‘SWOT-like data’. There are three current types of SWOT-like data: 1) AirSWOT, which is an airborne Ka-band inSAR currently limited to five rivers globally, 2) simulated rivers that mimic the type of data SWOT will provide, and 3) the SWOT simulator, which introduces measurement errors to these simulated rivers. Because we are principally interested in algorithm performance, we limit our validation setup to simulated rivers in order to benchmark across as many rivers as possible. These simulated rivers are simply reach-averaged hydraulic model outputs where the water surface heights and widths are labelled as ‘RS observations’ and are used as the sole inputs to the BIKER algorithm. Simulated rivers mimic perfect measurement conditions (i.e. no measurement errors). Here, we use the 22 rivers archived by @frassonCompilationHydraulicModels2019a and used by @rodriguezObservingRiversVarying2020 to develop reach-averaged Saint Vernant equations for the SWOT mission. We omitted the Arial Khan river in Bangladesh due to known problems with that hydraulic model **is that sufficient to say?** and we further sampled the models for only the observations every 11 days to mimic the average SWOT overpass frequency (this also approximately aligns the SWOT observations with the bi-weekly $CO_2$ data used later in this study- Section 2.4). All told, this yielded 503 sets of SWOT observations to validate BIKER with.

We also introduced measurement error into these simulated SWOT observations to assess BIKER's performance degradation due to errors inherent in SWOT measurements. Error in SWOT measurements will come from both the error tolerances intrinsic in the satellite data product as well as radar layover error. Here, we assume errors in river width are negligible [@durandHowWillRadar2020a] and solely focus on measurement errors in river height and slope. Layover error is the phenomenon when radar returns from different places arrive at the sensor at the same time, leading to taller landscape features appearing closer to the sensor than shorter landscape features that are the same horizontal distance from the sensor. This will occur when the landscape slope is steeper than the radar's incidence angle [@durandHowWillRadar2020a] and so generally occurs in areas of high topographic variation [@reesPhysicalPrinciplesRemote2013].

To explore the effect of SWOT measurement errors on BIKER, the SWOT simulator mentioned above is the ideal tool. However, it is cumbersome to run on many rivers **is this sufficient?** and so here we coarsely model measurement error using the results from a recent global anlaysis of SWOT water surface slope/height errors [@durandHowWillRadar2020a]. They developed a first-principles model for layover error from SWOT observations and, using global hydrography, quantified SWOT measurement error from both error tolerances and radar layover for 220,924 river reaches. The 68th percentiles of these reach-scale errors were 1.7cm/km and 10.4cm for water surface slope and height, respectively. To apply this model to our 22 rivers, we generate new SWOT observations by sampling every observation from normal distributions following equations 7 and 8 for node *i* and timestep *t*. Note that because BIKER ingests *dA*, rather than *H*, reach-scale height errors were converted to *dA* errors again following @durandHowWillRadar2020a and our rectangular channel assumptions: $\sigma_{dA} \approx W\sigma_H\sqrt{2}$. Note that equation 7 is also used to define slope measurement error in the MC analysis (section 2.2).

$$S_{err_{i,t}} \sim normal(S_{obs_{i,t}}, 0.000017^{2}) \mathbf{(7)}$$

$$dA_{err_{i,t}} \sim normal(dA_{obs_{i,t}}, \left[W_{obs_{i,t}}*\sqrt{2}*0.104\right]^{2}) \mathbf{(8)}$$

Regardless of the validation data used, we do not have observed $k_{600}$ data for these rivers, and to our knowledge no field dataset of $k_{600}$ exists in the type of temporal and spatial frequency that SWOT (and therefore the BIKER algorithm) provides. Further, we are principally interested in our ability to reproduce upscaled $k_{600}$ from RS observations and are less concerend with the actual accuracy of the upscaling model itself [which can be validated using existing datasets like previously done by @raymondScalingGasTransfer2012a; @ulsethDistinctAirWater2019a]. Therefore, we take the best performing upscaling model from sections 2.1 and 3.1 and use that to calculate the ‘observed’ $k_{600}$ that BIKER is validated against. This is done using equation 9, where $V_{obs} [L/T]$ is observed river discharge divided by observed channel area. With this setup, we can directly explore BIKER's ability to infer observed *V* and $k_{600}$ from river width and height alone. It also means that, for a fair validation scheme, $\sigma_{upscale}^2$ must be set to reflect only error from Manning’s equation (and not $a_i$ and $b_i$). Thus, $\sigma_{upscale}^2$ is set to 0.25 for this validation. However in practice, it should reflect the total uncertainty calculated in section 2.2 from all three sources.

$$k_{600, obs}=a_i(gS_hV_{obs})^{b_i} \mathbf{(9)}$$

Validation metrics take two forms (and are detailed in Table 1). To validate across all rivers and timesteps, we used the coefficient of determination $r^2$ and the root mean square error RMSE. Four normalized metrics were used for by-river validation: RRMSE and NRMSE are normalized root mean square errors that have been normalized by the observed value and the mean observed value (respectively) to compare across rivers. rBIAS is a measure of prediction bias that is normalized by the mean observed value to compare across rivers. KGE is a standard metric used in streamflow prediction with an intuitive basis: a value greater than 0 is often interpreted as a useful prediction in ungauged settings, and a value greater than -0.41 indicates a model outperforms a uniform prediction of the mean observed value [@knobenTechnicalNoteInherent2019a].

@@

###2.4 Upscaling to CO2 evasion and bulk carbon efflux
It is one thing to to predict $k_{600}$, but researchers are usually more interested in the gas fluxes from rivers and ultimately the bulk carbon efflux from river to atmosphere. Therefore, we explored 1) the ability of BIKER to reproduce hypothetical carbon dioxide evasion fluxes ($FCO_2$ $[M/L^2T]$) from these 22 rivers, and 2) the sensitivity of the estimated bulk carbon efflux $[M/T]$ via $CO_2$ evasion to the method used to estimate reach averaged flow velocity.

@@

###3.1 Validating the $k_{600}$ upscaling model
Figure 3 plots the validations for the five upscaling functions tested. Both the original @ulsethDistinctAirWater2019a model and our RS-able extension perform the best on this independent set of validation data, with near identical performance. The RS-able @ulsethDistinctAirWater2019a model achieves an $r^2$ of `r results_3_1[results_3_1$Model == 'RS-able',]$r2` and an RMSE of `r results_3_1[results_3_1$Model == 'RS-able',]$rmse` m/day, which is just marginally better than the original @ulsethDistinctAirWater2019a model (`r results_3_1[results_3_1$Model == 'Ulseth',]$r2` and `r results_3_1[results_3_1$Model == 'Ulseth',]$rmse` m/day, respectively). The two power-law based @raymondScalingGasTransfer2012a models (equations 4 and 3 in that study) achieve slightly worse performance, with $r^2$ in the 0.70s and their RMSE scores slightly higher (`r results_3_1[results_3_1$Model == 'Raymond_2',]$rmse` and `r results_3_1[results_3_1$Model == 'Raymond_3',]$rmse` m/day, respectively). @raymondScalingGasTransfer2012a equation 5, which is linear in structure, worked well on their dataset but fails to capture the non-linear relationship evident in this expanded and more geomorphically diverse dataset. In fact, it predicts negative $k_{600}$ values and error metrics cannot be calculated because the prediction residuals are not normally distributed. **SHOULD I INCLUDE THIS MODEL or not mention it??** Because we successfully reproduced the @ulsethDistinctAirWater2019a model, the rest of this manuscript deals exclusively with this RS-able variant of the model.

>>>>

##2 Methods
To build BIKER, we join a process-based model for $k_{o2}$ with a McFLI framework for inverting SWOT measurements (section 2.1). Following the description of that process, we describe the validation (section 2.3) and the workflow for comparing estimated bulk carbon effluxes from a suite of hydraulic geometry models (section 2.4). A flowchart detailing the entire study is given as Figure 1.

![Figure 1. Flowchart of methodology used in this study. We begin by deriving a process-based model for ko2 that uses only river width and slope to assign its parameters (section 2.1). Then, we implement this model within a McFLI ('BIKER') to estimate ko2 solely from river width and water surface slope (section 2.2). Finally, we couple BIKER with in situ CO2 data to compare BIKER-estimated gas fluxes with established in situ methods (section 2.4). See section 2.3 for the validation setup.](cache/flowchart.jpg)

###2.1 Developing BIKER
To use the McFLI framework, a physical model for the parameter of interest is necessary. In this case we need a physical model for *k*. First, we detail which model for *k* we chose to implement, and then we explain how the implementation works.

There is a long tradition of water quality engineers and biogeochemists developing predictive models for different forms of *k*, going back to at least the 1950s [@wangPhysicallyBasedScaling2021]. The goals of these models vary, from predicting oxygen reaeration coefficients to predicting $CO_2$ gas transfer velocities. Most of these models operate under the basic assumption that river channel hydraulics govern gas exchange and so simple statisical relationships can be generated between forms of *k* and hydraulic properties. However, many of them are not rooted in process-based mass transport theory and rather are simply empirical relations. This means that many of these models exist at the mercy of the limited training data available to fit said models and are highly sensitive to which data are used.

Recently, many of these models were rexplored in rivers and streams by @wangPhysicallyBasedScaling2021. They significantly expanded training datasets of field-measured $k_{o_2}$ (*k* for dissolved oxygen DO at 20 degrees Celsius) by using inverse modeling of DO timeseries. This apprroach was briefly mentioned in the introduction of this manuscript. These methods calibrate $k_{o_2}$, along with respiration and primary production, to high-fidelity, in situ datasets of DO with the goal of minimizing the difference between the observed DO and the mass-balance DO [e.g. @applingOvercomingEquifinalityLeveraging2018; @chapraDeltaMethodEstimating1991]. @wangPhysicallyBasedScaling2021 used this technique at 35 rivers across the United States and showed that $k_{o_2}$ models deduced from the classic, process-based 'thin-film' and 'surface renewal' theories yield approximately identical parameters when fit to field measurements or this new dataset of simulated $k_{o_2}$. This suggests that there are uniform scaling realtionships between certain hydraulic properties and $k_{o_2}$ that exist regardless of the training data used. See [@hallGasExchangeStreams2020 and @wangPhysicallyBasedScaling2021] for more in-depth explanations of the thin-film and surface renewal theories with respect to rivers. These are the standard models for understanding gas exchange in many aquatic environments from lakes to esutraries to rivers. The better fit, and more parsimonious, of these two models is reprinted as equations 2 and 3 where $U_\star$ is the shear velocity of the river channel [L/T] ($U_\star = \sqrt{gR_hS_h}$). Equation 2 was fit to field measurements of $k_{o_2}$ while equation 3 was fit to the simulated $k_{o_2}$ for 35 rivers.
$$k_{o2} = 47.67U_\star \mathbf{(2)}$$
$$k_{o2} = 49.71U_\star \mathbf{(3)}$$

This model was implemented in BIKER (using 48 as the approximate parameter across equations 2 and 3) for a few reasons. First, the analysis performed by [@wangPhysicallyBasedScaling2021] provides compelling evidence that a statistical parameter of approximately 48 is uniform across rivers and streams, regardless of the training data used. This is encouraging for use within BIKER as BIKER is specifically designed to be as river-agnostic as possible. Second, it yields a simple linear relationship across all rivers, regardless of their size or steepeness. This is not necessairly true of *k* models based on the turbulent dissipation rate *eD* ($eD=gS_hV$), which is another common approach [@hallGasExchangeStreams2020; e.g. @raymondScalingGasTransfer2012a; @oconnorMechanismReaerationNatural1958; @tsivoglouTracerMeasurementReaeration1976], where recent work has shown that *k* does not scale with $eD$ via the same statistical parameters across all rivers [@ulsethDistinctAirWater2019a]. We further explored the relationship presented in that paper and found that the $k~eD$ relationship fundamentally breaks down in rivers with low *eD*, which corresponds to nearly all rivers that SWOT will observe (**supp figure**). Therefore, an $eD$ based predictive model is of little use in SWOT rivers. Third, we sought to explicitly predict the gas transfer velocity for DO, as its use extends beyond constraining GHG fluxes from rivers. For example **stuff here**. To convert from $k_o2$ to *k* for any other GHG (as will be done later in this manuscript to calculate $CO_2$ evasion fluxes), we use the Schmidt number to convert between water temperatures and gases of interest following equation 4. 530 is the Schmidt number for oxygen at 20 degrees Celsius and *Sc* is the Schmidt number at one's desired temperature for some dissolved gas.
$$k_{GHG, temp}=k_{02}*(Sc/530)^{-1/2} \mathbf{(4)}$$

We stress that most other predictive models for *k* could be implemented within BIKER. However, our goal of using BIKER on any river that is observable by SWOT necessitates that we pick an equation that is applicable across as many rivers as possible. Future work should look at using a river-specific *k* equation to 'calibrate' BIKER to a specific river of interest.

With a predictive model for $k_{o2}$ chosen, we now implement it within a McFLI framework. The approach used here is strongly informed by the @hagemannBAMBayesianAMHGManning2017a McFLI algorithm for ungauged RSQ, further explored in more recent work [@brinkerhoffConstrainingRemoteRiver2020]. These papers conceptualize discharge as a Bayesian remote sensing problem, which we largely follow here to conceptualize $k_{o2}$ as a Bayesian remote sensing problem that can be solved using SWOT data.

First, the model needs to be rewritten as a function of SWOT-observables river width *W* and height (and thus water surface slope $S_h$). SWOT will measure these at approximately 200m intervals in pre-defined mass-conserved river reaches (**What is the PDD?**; SWORD). To do this, we follow @hagemannBAMBayesianAMHGManning2017a to formulate depth via river width and area. Area is further split into the SWOT-observable portion *dA* and the unobservable portion $A_0$ following @durandEstimatingReachaveragedDischarge2014a. $A_0$ is the median channel area and *dA* is the change in channel area. We estimate *dA* by assuming a rectangular river channel so that $dA \approx \delta H*W$ and is also solely a function of SWOT-observables *W* and $S_h$.

@@

$$k_{o2} = 48*U_\star=48\sqrt{gR_hS_h}=48g^{1/2}S_h^{1/2}(\frac{A_0+dA}{W})^{1/2} \mathbf{(5)}$$
Next, we rework equation 5 into a Bayesian sampling model to have all of the observations (*W* and $S_h$) on the left-hand side sampled from all of the unknown model parameters ($A_0$ and $k_{o_2}$) on the right-hand side. This is equation 6, where *x* are SWOT observables given $\Theta$, a set of statistical parameters. Note that in this formal likelihood specification both sides of equation 6 are log-transformed (natural log) to produce the normal sampling model presented here, however they are written in equation 6 without that transformation for succinctness. The $\sigma_{k_{o2}}^2$ parameter refers to the uncertainty inherent in equation 6's estimates. This will be explained in detail below.

$$f(x|\Theta) = \left[48^{-1}(\frac{W}{gS_h})^{1/2}\right] \sim normal\left(\left[k_{o2}^{-1}(A_0+dA)^{1/2}\right], \sigma^2_{k_{o2}}\right)  \mathbf{(6)}$$

Equation six necessitates that Bayesian priors be assigned to parameters $A_0$ and $k_{o_2}$. Bayesian priors formalize the a priori estimates (and uncertainties) for the non-remotely-sensed parameters. To state that more intuitively for these physical quantities, priors represent our prior river knowledge of what $A_0$ and $k_{o_2}$ probably are for some river given that they cannot be remotely sensed. All priors are formalized within the model as truncated normal distributions of the log-transformed terms such that $log(X) \sim N(\mu, \sigma^2)$ for $\lambda < X < \epsilon$, using prior hyperparameters mean ($\mu$), standard deviation ($\sigma$), and upper ($\epsilon$) and lower bounds ($\lambda$) for parameter *X*. In order to avoid relying on in situ information, we assign prior hyperparameters using SWOT data only. $A_0$ prior hyperparameters were assigned following the method developed by @brinkerhoffConstrainingRemoteRiver2020. They developed a set of river channel prior hyperparameters for McFLI algorithms that are entirely RS-able and reflect differential channel hydraulics as a function of river geomorphology. They used an extensive database of field measurements and machine learning to identify patterns that associate the river width with river geomorphic properties, so these non-remotely sensible properties may be assigned to rivers using only the existing remotely sensed data. This leaves the $k_{o_2}$ hyperparameters to be defined. We assigned the $k_{o_2}$ prior hyperparameters using a simple slope regression model trained on the @raymondScalingGasTransfer2012a dataset to assign $\mu$ (equation 7). $\sigma^2$ was set to the standard error of equation 7: 0.77. $\lambda$ and $\epsilon$ were set to log(0.0001) and log(60), respectively.

$$\mu_{k_{o_2}}=3.3492 +0.3661ln(S_h)   \mathbf{(7)}$$

**FIGURE OUT Finally, we estimate $σ_{k600}^2$ using Monte Carlo (MC) methods to approximate total uncertainty from equation 3. Uncertainty in equation 3 can stem from 1) $a_i$ and $b_i$ parameter uncertainty, 2) error in Manning’s approximation of the average flow velocity, and 3) measurement error. Total equation 3 uncertainty is a function of all three uncertainity sources, so we need to propogate the three uncertanties through the algebra of equation 3. Here, we use MC simulations to do this. MC simulations repeately sample from parameter distributions defined by their individual expectations and uncertainities to produce a distribution of model estimates from which an overall uncertainity term can be extracted. Here, we run 5,000 unique MC simulations on 5,000 sets of field measurements of river channel hydraulics from @brinkerhoffReconcilingAtaStationAtManyStations2019 and use the median uncertainity term across those 5,000 distributions as $\sigma_{k600}$. Each MC simulation is itself 10,000 runs, sampling from the normal distributions for $a_i$, $b_i$, the log-transformed flow velocity, and $S_h$. Uncertainty for velocity followed @hagemannBAMBayesianAMHGManning2017a, who estimated ln-transformed Manning’s equation error in discharge to be 0.25 (we assume that the 0.25 value is due soley from Manning's equation and not measurements of cross-sectional area in order to use here as the velocity error). The slope-error model that we used is explained in detail in Section 2.3 and relies on a recent analysis of expected errors in SWOT $S_h$ measurements [@durandHowWillRadar2020a].**

With the sampling model (equation 4), prior distributions, and the $\sigma_{k_{o_2}}^2$ parameter described, a joint posterior distribution conditional on the SWOT data is obtained. To approximate this distribution, we use a Markov Chain Monte Carlo (MCMC) algorithm. Because it is written in the Stan probabilistic programming language, we use a Hamiltonian Monte Carlo which reduces computation time relative to other sampling algorithms [again following @hagemannBAMBayesianAMHGManning2017a].

###2.3 BIKER validation setup
We validated BIKER on 50 rivers using observed $R_h$ and observed $S_h$ First, we detail the 50 rivers, and then we detail how observed $k_{o_2}$ is calculated.

Because SWOT has yet to launch, it is standard practice to benchmark SWOT-related algorithms on SWOT-like data. There are three current types of SWOT-like data: 1) AirSWOT, which is an airborne Ka-band inSAR currently limited to five rivers globally, 2) simulated rivers that mimic the type of data SWOT will provide, and 3) the SWOT simulator, which introduces measurement errors to these simulated rivers. Because we are principally interested in algorithm performance, we limit our validation setup to simulated rivers in order to benchmark across as many rivers as possible. These simulated rivers are simply reach-averaged hydraulic model outputs where the water surface heights and widths are labelled as RS observations and are used as the sole inputs to the BIKER algorithm. Simulated rivers mimic perfect measurement conditions (i.e. no measurement errors). Here, we use the 50 rivers developed by @frassonExploringFactorsControlling and @durandIntercomparisonRemoteSensing2016b. These are the two benchmarking studies that have explored RSQ algorithm performance for the SWOT mission. Consult both of those papers for maps of the river's locations, though the 50 rivers are spread across the United States, Canada, France, Italy, United Kingdom, India, and Bangladesh. We omitted the Arial Khan river in Bangladesh due to known problems with that hydraulic model **is that sufficient to say?**.

We also assess the influence of measurement error on BIKER's performance. Error in SWOT measurements will come from both the error tolerances intrinsic in the satellite data product as well as radar layover error and width measurement error. Layover error is the phenomenon when radar returns from different places arrive at the sensor at the same time, leading to taller landscape features appearing closer to the sensor than shorter landscape features that are the same horizontal distance from the sensor. This will occur when the landscape slope is steeper than the radar's incidence angle [@durandHowWillRadar2020a] and so generally occurs in areas of high topographic variation [@reesPhysicalPrinciplesRemote2013].

For river heights and slopes, we use the SWOT error model devolped by @durandHowWillRadar2020a and implemented on 16 of the rivers in this study by @frassonExploringFactorsControlling. Systematic height errors are assumed constant across the river but variable in time. Width errors were derived using a model built by @frassonExploringFactorsControlling based on SWOT-simulator runs for the Sacramento and Po rivers. Width errors due to poor water classification are ignored here. Please consult @frassonExploringFactorsControlling for the specifics of how these realistic errors were incorporated into the hydraulic models. Ultimately, we reran BIKER on these 16 rivers with the SWOT error model corrupting the SWOT-observables *W* and *S_h*.

Regardless of the validation data or SWOT error budget used, we do not have observed $k_{o_2}$ data for these rivers, and to our knowledge no field dataset of $k_{o_2}$ exists in the type of temporal and spatial frequency that SWOT (and therefore the BIKER algorithm) provides. Further, we are principally interested in BIKER's ability to reproduce $k_{o_2}$ from SWOT observations and are less concerned with the actual accuracy of the predictive model itself [which can be validated using existing datasets like previously done by @wangPhysicallyBasedScaling2021]. Therefore, we take the model outlined in section 2.1 and use that to calculate the observed $k_{o_2}$ that BIKER is validated against. This is done using equation 8, where $R_{h, obs} [L/T]$ is observed river channel area divided by observed channel width. With this setup, we can directly explore BIKER's ability to infer observed $U_\star$ and $k_{o_2}$ from *W* and $S_h$ alone. It also means that, for a fair validation scheme, $\sigma_{upscale}^2$ must be set to reflect only error from Manning's equation (and not the parameter 48 in equation 8). Thus, $\sigma_{upscale}^2$ is set to 0.25 for this validation. However in practice, it should reflect the total uncertainty calculated in section 2.2 from all three sources.

$$k_{o_2, obs}=48\sqrt{gS_hR_{h, obs}} \mathbf{(8)}$$

Validation metrics take two forms (and are detailed in Table 1). To validate across all rivers and timesteps, we used the coefficient of determination $r^2$ and the root mean square error RMSE. Four normalized metrics were used for by-river validation: RRMSE and NRMSE are normalized root mean square errors that have been normalized by the observed value and the mean observed value (respectively). rBIAS is a measure of prediction bias that is normalized by the mean observed value. KGE is a standard metric used in streamflow prediction with an intuitive basis: a value greater than 0 is often interpreted as a useful prediction in ungauged settings, and a value greater than -0.41 indicates a model outperforms a uniform prediction of the mean observed value [@knobenTechnicalNoteInherent2019a].

@@

###2.4 Upscaling to CO2 evasion and bulk carbon efflux
It is one thing to to predict $k_{o_2}$, but researchers are usually more interested in the gas fluxes from rivers and ultimately the bulk carbon efflux from river to atmosphere. Therefore, we explored 1) the ability of BIKER to reproduce hypothetical carbon dioxide evasion fluxes ($FCO_2$ $[M/L^2T]$) from these 22 rivers, and 2) the sensitivity of the estimated bulk carbon efflux $[M/T]$ via $CO_2$ evasion to the method used to estimate reach averaged flow velocity.

@@
